Scalable analysis of genetic variation data in Python

### 2024-01-10 paper.tex

\title{Scalable statistical genetic data analysis in Python}

@@

%%%
%%% Previous abstract
%%%
% The PyData ecosystem is an umbrella term covering Python packages based on a
% broad range of modern techniques, such as chunk-compressed columnar data
% storage, just-in-time compilation of numerical code, and scaling of
% calculations across clusters of computers. Together, these technologies have
% been successfully applied in scientific applications using data at the petabyte
% scale. These technologies, and the many benefits that they provide, however,
% have not been successfully applied in the field of genomics, which is currently
% making the transition to working at petabyte scale. We present sgkit, a Python
% package designed to bring the benefits of the PyData ecosystem to genomics,
% allowing users to efficiently analyse large-scale data using familiar tools. We
% discuss the underlying design principles of these technologies and illustrate
% their suitability in genetics and genomics applications, via examples on
% large-scale datasets such as UK Biobank.

\begin{abstract} The Abstract (250 words maximum) should be structured to
include the following details:
\textbf{Background}, the context and purpose of
the study;
\textbf{Results}, the main findings;
\textbf{Conclusions}, brief
summary and potential implications. Please minimize the use of abbreviations
and do not cite references in the abstract.
\end{abstract}

>>>>

\title{Scalable analysis of genetic variation data in Python}

@@

% The Abstract (250 words maximum) should be structured to
% include the following details:
% \textbf{Background}, the context and purpose of
% the study;
% \textbf{Results}, the main findings;
% \textbf{Conclusions}, brief
% summary and potential implications. Please minimize the use of abbreviations
% and do not cite references in the abstract.
% The Abstract (250 words maximum) should be structured to
% include the following details:
% \textbf{Background}, the context and purpose of % the study;
% \textbf{Results}, the main findings;
% \textbf{Conclusions}, brief summary and potential implications.

%% NOTE: this is much too long currently, but keeping for now so we 
% can see which stuff migrates to the intro
\begin{abstract}
\textbf{Background:}
Vast quantities of genetic variation data have been collected to 
address fundamental scientific questions across multiple disciplines.
Genetic variation datasets are usually modelled as a ``genotype matrix'',
where rows denote positions on the genome where variation occurs and 
columns represent the state of sequenced samples at that position.
The largest dataset available today, the UK Biobank, 
has a genotype matrix of approximately 1,000,000,000 rows and 500,000 columns,
along with metadata and quality control information of similar dimension.
Processing such matrices on a single computer is not feasible,
and computations must be distributed to complete in a reasonable timeframe.
Classical methods in genomics are inherently single-computer focused, 
oriented around either streaming data through Unix pipelines or 
storing the entire genotype matrix in memory.
The standard file format for genetic variation data (VCF: Variant Call Format)
is also fundamentally unsuited to distributed computation and is a severe 
bottleneck at the Biobank scale.
Researchers either perform ad-hoc distribution by splitting across VCF files
and dispatching jobs using workflow managers, 
or convert data into the undocumented and/or proprietary formats 
used by the few specialised tools that do support automatic distribution
of calculations on a genotype matrix.

\textbf{Results:} We present sgkit, a scalable toolkit for processing genetic 
variation data built on widely-used technologies from the Python data-science
ecosystem. A key part of this approach is to use Zarr, a cloud-native
format for storing multi-dimensional numerical data, widely used in the
across the sciences to store petabyte-scale datasets, and with 
multiple implementations.  We show that data stored in this format has 
comparable compression levels to specialised methods to compress the 
genotype matrix. We show that performing calculations on the genotype
matrix on a single computer is much more efficient at scale using sgkit than standard 
approaches based on VCF, and is comparable with state-of-the-art methods
using compiled languages. Moreover, sgkit can also natively distribute
computation over a cluster using Dask, with no changes required to client code.
We show what sgkit is competitive with Hail (the most popular package for distributed
analysis of genetic variation data) in terms of performance, but is 
much more flexible and extensible. We demonstrate the 
current capabilities of sgkit to perform dataset quality control,
and applications in statistical, population, quantitative and phylo genetics.

\textbf{Conclusions:} 
Sgkit enables the interactive analysis of Biobank-scale genetic variation
datasets via familiar and powerful Python data science tools, boosting
researcher productivity. Leveraging these tools also substantially 
reduces software development effort for creating new methods.
The vcf-zarr specification may additionally provide a useful starting 
point for defining a standard computational format for genetic variation data.
Such a standard, based on the language-agnostic Zarr format, would enable a
diverse ecosystem of tools to efficiently access subsets of the data without
copying, and substantially reduce research costs.
\end{abstract}
